---
title: "Final Project Code Appendix"
output:
  html_document: default
---

## 0. Imports and Load Packages
```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(lmtest)
library(sandwich)
library(scales)
```

## 1. Read and clean data
```{r}
salary <- read.table(
  "salary.txt",
  header = TRUE,
  sep = "",
  stringsAsFactors = FALSE,
  na.strings = "NA"
)

salary <- salary %>%
  mutate(
    sex = factor(sex, levels = c("M", "F")),
    deg = factor(deg),
    field = factor(field),
    rank = factor(rank, levels = c("Assist", "Assoc", "Full")),
    admin = factor(admin, levels = c(0, 1)),
    startyr = as.integer(startyr),
    year = as.integer(year),
    yrdeg = as.integer(yrdeg),
    salary = as.numeric(salary),
    experience = year - startyr
  )

# checks for problematic values (ensured no negative experience)
# salary %>%
  # filter(experience < 0) %>%
  # print()

salary_clean <- salary %>%
  filter(experience >= 0, !is.na(rank))
```

## 2. Descriptive summaries
```{r}
salary_clean %>%
  count(sex, rank, field)

salary_clean %>%
  summarise(
    n_records = n(),
    n_people = n_distinct(id),
    min_year = min(year, na.rm = TRUE),
    max_year = max(year, na.rm = TRUE)
  ) %>%
  print()

# colour palette for sex used across all plots
sex_colours <- c("M" = "#4472C4", "F" = "#E05C5C")
```

## 3. Q1: Starting salary discrimination
```{r}
start_salary <- salary_clean %>%
  filter(year == startyr)

start_salary %>%
  summarise(
    n_records = n(),
    n_people = n_distinct(id)
  ) %>%
  print()

cat("\nRank distribution at hire:\n")
start_salary %>%
  count(rank, sex) %>%
  arrange(rank, sex)

cat("\nStarting salary summaries by sex:\n")
q1_summary <- start_salary %>%
  group_by(sex) %>%
  summarise(
    n = n(),
    mean_salary = mean(salary, na.rm = TRUE),
    median_salary = median(salary, na.rm = TRUE),
    sd_salary = sd(salary, na.rm = TRUE),
    iqr_salary = IQR(salary, na.rm = TRUE)
  )
print(q1_summary)

cat("\nStarting salary summaries by sex and field:\n")
start_salary %>%
  group_by(field, sex) %>%
  summarise(
    n = n(),
    mean_salary = mean(salary, na.rm = TRUE),
    median_salary = median(salary, na.rm = TRUE),
    .groups = "drop"
  )
```
```{r}
par(mfrow = c(1, 2))
hist(start_salary$salary, main = "Starting Salary Histogram", xlab = "Salary")
qqnorm(start_salary$salary, main = "Q-Q Plot: Starting Salary")
qqline(start_salary$salary, col = "red")
par(mfrow = c(1, 1))

ggplot(start_salary, aes(x = salary, fill = sex)) +
  geom_histogram(alpha = 0.6, bins = 25, position = "identity") +
  scale_fill_manual(values = sex_colours) +
  facet_wrap(~ sex, scales = "free_y") +
  labs(
    title = "Starting Salary Distribution by Sex",
    x = "Starting Salary",
    y = "Count"
  )

ggplot(start_salary, aes(x = sex, y = salary, fill = sex)) +
  geom_boxplot(alpha = 0.8) +
  scale_fill_manual(values = sex_colours) +
  labs(
    title = "Starting Salary by Sex",
    x = "Sex",
    y = "Starting Salary"
  )

ggplot(start_salary, aes(x = factor(startyr), y = salary, fill = sex)) +
  geom_boxplot(outlier.alpha = 0.3) +
  scale_fill_manual(values = sex_colours) +
  labs(
    title = "Starting Salary by Hire Year and Sex",
    x = "Hire Year",
    y = "Starting Salary"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
cat("\nQ1 Welch two-sample t-test:\n")
q1_ttest <- t.test(salary ~ sex, data = start_salary)
print(q1_ttest)

q1_lm <- lm(salary ~ sex + field + yrdeg + startyr, data = start_salary)
cat("\nQ1 linear regression on raw salary:\n")
summary(q1_lm)
```

```{r}
par(mfrow = c(2, 2))
plot(q1_lm)
par(mfrow = c(1, 1))

cat("\nQ1 regression with heteroskedasticity-robust SE:\n")
coeftest(q1_lm, vcov = vcovHC(q1_lm, type = "HC1"))

# Log salary model
start_salary <- start_salary %>%
  mutate(log_salary = log(salary))

q1_lm_log <- lm(log_salary ~ sex + field + yrdeg + startyr, data = start_salary)
cat("\nQ1 linear regression on log salary:\n")
summary(q1_lm_log)

cat("\nQ1 log-salary regression with robust SE:\n")
coeftest(q1_lm_log, vcov = vcovHC(q1_lm_log, type = "HC1"))

q1_log_coef <- coef(q1_lm_log)["sexF"]
cat("\nApproximate percent difference for women in starting salary:\n")
cat(round((exp(q1_log_coef) - 1) * 100, 2), "%\n")
```


## 4a. Q2: Wage difference at promotion
```{r}
promotion_data <- salary_clean %>%
  arrange(id, year) %>%
  group_by(id) %>%
  mutate(
    prev_rank = lag(rank),
    prev_salary = lag(salary),
    prev_year = lag(year)
  ) %>%
  ungroup() %>%
  filter(prev_rank == "Assoc", rank == "Full")

promotion_data %>%
  summarise(
    n_promotions = n(),
    n_people = n_distinct(id)
  ) %>%
  print()

promotion_data <- promotion_data %>%
  group_by(id) %>%
  slice(1) %>%
  ungroup()
```
```{r}
promotion_data <- promotion_data %>%
  mutate(
    raise_amount = salary - prev_salary,
    raise_pct = (salary - prev_salary) / prev_salary
  )

cat("\nPromotion raise summaries by sex:\n")
promotion_data %>%
  group_by(sex) %>%
  summarise(
    n = n(),
    mean_raise = mean(raise_amount, na.rm = TRUE),
    median_raise = median(raise_amount, na.rm = TRUE),
    mean_pct_raise = mean(raise_pct, na.rm = TRUE),
    median_pct_raise = median(raise_pct, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  print()
```

```{r}
ggplot(promotion_data, aes(x = sex, y = raise_amount, fill = sex)) +
  geom_boxplot(alpha = 0.8) +
  scale_fill_manual(values = sex_colours) +
  labs(
    title = "Salary Increase at Promotion (Associate to Full) by Sex",
    x = "Sex",
    y = "Dollar Raise at Promotion"
  )

ggplot(promotion_data, aes(x = sex, y = raise_pct, fill = sex)) +
  geom_boxplot(alpha = 0.8) +
  scale_fill_manual(values = sex_colours) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    title = "Percent Raise at Promotion by Sex",
    x = "Sex",
    y = "Percent Raise"
  )

ggplot(promotion_data, aes(x = raise_pct, fill = sex)) +
  geom_histogram(alpha = 0.6, bins = 20, position = "identity") +
  scale_fill_manual(values = sex_colours) +
  facet_wrap(~ sex) +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    title = "Distribution of Percent Raise at Promotion",
    x = "Percent Raise",
    y = "Count"
  )
```

```{r}
cat("\nQ2 Welch t-test for dollar raise:\n")
print(t.test(raise_amount ~ sex, data = promotion_data))

cat("\nQ2 Welch t-test for percent raise:\n")
print(t.test(raise_pct ~ sex, data = promotion_data))

cat("\nQ2 Wilcoxon test for percent raise:\n")
print(wilcox.test(raise_pct ~ sex, data = promotion_data))
```

```{r}
q2_lm_raise <- lm(
  raise_amount ~ sex + field + yrdeg + prev_salary,
  data = promotion_data
)
cat("\nQ2 regression for dollar raise:\n")
summary(q2_lm_raise)

cat("\nQ2 regression for dollar raise with robust SE:\n")
coeftest(q2_lm_raise, vcov = vcovHC(q2_lm_raise, type = "HC1"))

q2_lm_pct <- lm(
  raise_pct ~ sex + field + yrdeg + prev_salary,
  data = promotion_data
)
cat("\nQ2 regression for percent raise:\n")
summary(q2_lm_pct)

cat("\nQ2 regression for percent raise with robust SE:\n")
coeftest(q2_lm_pct, vcov = vcovHC(q2_lm_pct, type = "HC1"))
```

### 4b. Q2: Promotion Rates and Within-Rank Wage Gaps
This section investigates whether there are gender differences in promotion rates from Associate Professor to Full Professor, as well as differences in salary within academic ranks. Specifically, we examine (1) whether men and women are promoted to Full Professor at different rates, (2) how long faculty spend as Associate Professors before promotion, and (3) whether women earn systematically different salaries within the Associate and Full ranks after controlling for relevant characteristics.

#### Method

We first construct a **person-level dataset** restricted to faculty who have ever held the rank of Associate Professor. For each individual, we identify whether they were eventually promoted to Full Professor, the year they first became an Associate Professor, and the year they were first promoted to Full Professor (if applicable). We also compute the number of years spent as an Associate Professor before promotion. Additionally, we record salary at Associate entry, the last Associate salary, and the salary at entry into the Full Professor rank.

This person-level dataset allows us to compare promotion outcomes and salary changes across individuals.

```{r}
```{r}
# build person-level dataset restricted to ever-Associate faculty
ever_assoc_ids <- salary_clean %>%
  group_by(id) %>%
  filter(any(rank == "Assoc")) %>%
  pull(id) %>%
  unique()

person_level <- salary_clean %>%
  filter(id %in% ever_assoc_ids) %>%
  arrange(id, year) %>%
  group_by(id) %>%
  summarise(
    sex             = first(sex),
    deg             = first(deg),
    field           = first(field),
    yrdeg           = first(yrdeg),
    promoted        = as.integer(any(rank == "Full")),
    yr_first_assoc  = min(year[rank == "Assoc"], na.rm = TRUE),
    yr_first_full   = ifelse(any(rank == "Full"),
                             min(year[rank == "Full"], na.rm = TRUE), NA_real_),
    yrs_as_assoc    = ifelse(any(rank == "Full"),
                             yr_first_full - yr_first_assoc, NA_real_),
    sal_assoc_entry = salary[rank == "Assoc"][which.min(year[rank == "Assoc"])],
    sal_assoc_last  = salary[rank == "Assoc"][which.max(year[rank == "Assoc"])],
    sal_full_entry  = ifelse(any(rank == "Full"),
                             salary[rank == "Full"][which.min(year[rank == "Full"])],
                             NA_real_),
    .groups = "drop"
  )
```
#### Promotion Rates by Sex

We first compare promotion rates between men and women by computing the proportion of Associate Professors who were eventually promoted to Full Professor by the end of the observation period.
```{r}
cat("Promotion rates by sex:\n")
person_level %>%
  group_by(sex) %>%
  summarise(
    n          = n(),
    n_promoted = sum(promoted),
    pct        = round(100 * mean(promoted), 1),
    .groups = "drop"
  ) %>%
  print()

ct <- table(Sex = person_level$sex, Promoted = person_level$promoted)
colnames(ct) <- c("Not Promoted", "Promoted")
cat("\nContingency table:\n")
print(ct)
cat("\nRow proportions:\n")
print(round(prop.table(ct, margin = 1), 3))

cat("\nMedian time as Associate before promotion:\n")
person_level %>%
  filter(promoted == 1) %>%
  group_by(sex) %>%
  summarise(median_yrs = median(yrs_as_assoc, na.rm = TRUE), .groups = "drop") %>%
  print()
```
#### Observations

The results show that 59.9% of male Associate Professors were promoted to Full Professor, compared to 42.0% of female Associate Professors. The contingency table indicates that 447 of 746 men were promoted, while only 100 of 238 women were promoted.

Examining row proportions shows that about 60% of men were promoted compared to 42% of women, suggesting a substantial gender gap in promotion outcomes.

Among those who were promoted, the median time spent as an Associate Professor was 5 years for men and 6 years for women, indicating that women tend to remain in the Associate rank slightly longer before promotion.


#### Visualization of Promotion and Salary Patterns

To better understand these patterns, we visualize promotion rates, entry salaries at the Associate rank, salary gains upon promotion, and the distribution of time spent as an Associate Professor.
```{r}
# promotion rate bar chart
ggplot(person_level %>% group_by(sex) %>% summarise(pct = mean(promoted)),
       aes(x = sex, y = pct, fill = sex)) +
  geom_col(width = 0.5, alpha = 0.8) +
  scale_fill_manual(values = sex_colours) +
  scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0, 0.75)) +
  labs(title = "Promotion Rate (Associate to Full) by Sex",
       x = "Sex", y = "Proportion promoted by 1995")

# salary at Associate entry by sex
ggplot(person_level, aes(x = sex, y = sal_assoc_entry, fill = sex)) +
  geom_boxplot(alpha = 0.8) +
  scale_fill_manual(values = sex_colours) +
  labs(title = "Salary at Associate Entry by Sex",
       x = "Sex", y = "Monthly salary ($)")

# salary gain at promotion by sex
ggplot(person_level %>% filter(promoted == 1) %>%
         mutate(sal_gain = sal_full_entry - sal_assoc_last),
       aes(x = sex, y = sal_gain, fill = sex)) +
  geom_boxplot(alpha = 0.8) +
  scale_fill_manual(values = sex_colours) + 
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  labs(title = "Salary Gain at Promotion by Sex",
       x = "Sex", y = "Salary gain ($)")

# time as Associate before promotion
ggplot(person_level %>% filter(promoted == 1),
       aes(x = yrs_as_assoc, fill = sex)) +
  geom_histogram(alpha = 0.6, bins = 15, position = "identity") +
  scale_fill_manual(values = sex_colours) +
  labs(title = "Years as Associate Before Promotion",
       x = "Years", y = "Count")
```
#### Statistical Test for Differences in Promotion Rates
To formally test whether the difference in promotion rates between men and women is statistically significant, we conduct a two-sample z-test for the difference in proportions. This test evaluates the null hypothesis that the promotion probability is the same for men and women.

```{r}
# two-sample z-test for difference in promotion proportions
nF <- sum(person_level$sex == "F")
nM <- sum(person_level$sex == "M")
kF <- sum(person_level$sex == "F" & person_level$promoted == 1)
kM <- sum(person_level$sex == "M" & person_level$promoted == 1)
pF <- kF / nF
pM <- kM / nM

p_pool <- (kF + kM) / (nF + nM)
se_z   <- sqrt(p_pool * (1 - p_pool) * (1/nF + 1/nM))
z_stat <- (pF - pM) / se_z
p_val  <- 2 * pnorm(-abs(z_stat))

se_diff <- sqrt(pF*(1-pF)/nF + pM*(1-pM)/nM)
ci_lo   <- (pF - pM) - 1.96 * se_diff
ci_hi   <- (pF - pM) + 1.96 * se_diff

cat(sprintf("pF = %.4f  (%d/%d)\n", pF, kF, nF))
cat(sprintf("pM = %.4f  (%d/%d)\n", pM, kM, nM))
cat(sprintf("Difference pF - pM = %+.4f\n", pF - pM))
cat(sprintf("95%% CI: (%.4f, %.4f)\n", ci_lo, ci_hi))
cat(sprintf("z = %.3f,  p = %.4f\n", z_stat, p_val))

cat("\nEquivalent chi-square test:\n")
print(chisq.test(ct, correct = FALSE))
```

#### Observations

The promotion rate for women is 0.420, while the promotion rate for men is 0.599, producing a difference of −0.179. The 95% confidence interval for this difference is (−0.251, −0.107), which does not include zero.

The z-test statistic is −4.84, with a p-value less than 0.001, indicating that the difference in promotion rates between men and women is statistically significant. The equivalent chi-square test yields the same conclusion.

These results provide strong evidence that women were less likely to be promoted from Associate to Full Professor during the sample period.

#### Within-Rank Wage Gap Analysis

Next, we examine whether salary differences persist within academic ranks after controlling for relevant characteristics. For each rank (Associate and Full Professor), we estimate a log-linear regression model where the dependent variable is the logarithm of salary.

The model controls for:

sex

year of degree

calendar year

administrative roles

years of experience

academic field

degree type

Using log salary allows the coefficient on the gender indicator to be interpreted approximately as a percentage difference in salary.
```{r}
# log-salary regression within rank to estimate adjusted wage gap by sex
for (rnk in c("Assoc", "Full")) {
  sub <- salary_clean %>%
    filter(rank == rnk, salary > 0) %>%
    mutate(log_salary = log(salary), years_exp = year - startyr)

  fit <- lm(
    log_salary ~ sex + yrdeg + year + admin + years_exp + field + deg,
    data = sub
  )

  cat(sprintf("\nQ2 log-salary regression — %s rank (n = %d):\n", rnk, nrow(sub)))
  summary(fit)

  cat("\nWith robust SE:\n")
  print(coeftest(fit, vcov = vcovHC(fit, type = "HC1")))

  coef_f <- coef(fit)["sexF"]
  cat(sprintf("Approximate percent difference for women in salary: %.1f%%\n",
              (exp(coef_f) - 1) * 100))
}
```
#### Results

The regression results indicate a statistically significant negative coefficient on the female indicator in both ranks.

Among Associate Professors, the coefficient on sexF corresponds to an approximate salary difference of −5.3%, meaning that women earn about 5.3% less than men on average after controlling for experience, year effects, field, degree type, and administrative roles.

Among Full Professors, the estimated difference is −5.5%, indicating that women continue to earn roughly 5–6% less than men even after reaching the highest academic rank and after accounting for observable characteristics.

#### Interpretation

Taken together, these results suggest two important patterns. First, women were less likely to be promoted from Associate to Full Professor during the study period. Second, conditional on rank and other observed characteristics, women earned modestly lower salaries than men within both the Associate and Full Professor ranks. These findings suggest that gender disparities may arise both through promotion dynamics and within-rank wage differences.

## 5. Q3: Overall salary discrimination
```{r}
# avoid longitudinal methods by using one row per faculty (latest salary)
latest_salary <- salary_clean %>%
  arrange(id, year) %>%
  group_by(id) %>%
  slice_tail(n = 1) %>%
  ungroup() %>%
  mutate(
    log_salary = log(salary),
    experience = year - startyr
  )

latest_salary %>%
  summarise(
    n_people = n(),
    min_year = min(year, na.rm = TRUE),
    max_year = max(year, na.rm = TRUE)
  ) %>%
  print()
```

```{r}
cat("\nLatest salary summaries by sex and rank:\n")
latest_salary %>%
  group_by(rank, sex) %>%
  summarise(
    n = n(),
    mean_salary = mean(salary, na.rm = TRUE),
    median_salary = median(salary, na.rm = TRUE),
    sd_salary = sd(salary, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(latest_salary, aes(x = sex, y = salary, fill = sex)) +
  geom_boxplot(alpha = 0.8) +
  scale_fill_manual(values = sex_colours) +
  facet_wrap(~ rank, scales = "free_y") +
  labs(
    title = "Latest Salary by Sex Within Rank",
    x = "Sex",
    y = "Salary"
  )

ggplot(latest_salary, aes(x = salary, fill = sex)) +
  geom_density(alpha = 0.4) +
  scale_fill_manual(values = sex_colours) + 
  facet_wrap(~ rank, scales = "free") +
  labs(
    title = "Latest Salary Density by Sex Within Rank",
    x = "Salary",
    y = "Density"
  )
```

```{r}
q3_lm <- lm(
  salary ~ sex + rank + field + experience + admin,
  data = latest_salary
)
cat("\nQ3 overall salary model (latest salary):\n")
summary(q3_lm)

cat("\nQ3 overall salary model with robust SE:\n")
coeftest(q3_lm, vcov = vcovHC(q3_lm, type = "HC1"))

# interaction model: does the salary gap differ by rank?
q3_interaction <- lm(
  salary ~ sex * rank + field + experience + admin,
  data = latest_salary
)
cat("\nQ3 salary model with sex x rank interaction:\n")
summary(q3_interaction)

cat("\nQ3 interaction model with robust SE:\n")
coeftest(q3_interaction, vcov = vcovHC(q3_interaction, type = "HC1"))

# log-salary
q3_lm_log <- lm(
  log_salary ~ sex + rank + field + experience + admin,
  data = latest_salary
)
cat("\nQ3 log-salary model:\n")
summary(q3_lm_log)

cat("\nQ3 log-salary model with robust SE:\n")
coeftest(q3_lm_log, vcov = vcovHC(q3_lm_log, type = "HC1"))

# coefs
q3_log_coef <- coef(q3_lm_log)["sexF"]
cat("\nApproximate percent difference for women in latest salary:\n")
cat(round((exp(q3_log_coef) - 1) * 100, 2), "%\n")
```





