---
title: "Question 1: Does Wage Discrimination Exist in Starting Salaries?"
author: "Ha Tien Nguyen"
date: today
format:
  html:
    toc: true
    code-fold: true
    self-contained: true
execute:
  warning: false
  message: false
---

**Research Question:** Does wage discrimination exist in the starting salaries of faculty members (i.e., salaries in the year hired)?

**Significance level:** α = 0.05 (two-sided tests throughout).

---

# 1. Setup

```{r setup}
library(ggplot2)
library(dplyr)
library(sandwich)
library(lmtest)
```

---

# 2. Load & Clean Data

```{r}
df <- read.table("salary.txt", header = TRUE, strip.white = TRUE)
df$sex   <- factor(df$sex,   levels = c("M", "F"))
df$field <- factor(df$field)

cat("Total number of rows: ", nrow(df), "rows\n")
```


## 2.6 Final Dataset

```{r}
cat("Final dataset:", nrow(df), "faculty members\n")
cat("  Female:", sum(df$sex == "F"), " | Male:", sum(df$sex == "M"), "\n")
cat("  Hire years:", min(df$startyr), "–", max(df$startyr), "\n")
cat("  Salary range: $", min(df$salary), "– $", max(df$salary), "\n")
```

---

# 3. Descriptive Statistics

## 3.1 Summary by Sex

```{r}
tapply(df$salary, df$sex, function(x) {
  round(c(n = length(x), mean = mean(x), median = median(x),
          sd = sd(x), IQR = IQR(x)), 2)
})
```

## 3.2 Boxplots by Sex

```{r}
ggplot(df, aes(x = sex, y = salary, fill = sex)) +
  geom_boxplot(alpha = 0.7) +
  scale_x_discrete(labels = c(M = "Male", F = "Female")) +
  scale_fill_manual(values = c(M = "#ff7f0e", F = "#1f77b4"), guide = "none") +
  labs(title = "Starting Salary by Sex", x = NULL, y = "Monthly Salary ($)") +
  theme_minimal()
```

## 3.3 Histograms by Sex

```{r}
ggplot(df, aes(x = salary, fill = sex)) +
  geom_histogram(bins = 30, colour = "white", alpha = 0.8) +
  facet_wrap(~sex, ncol = 1, scales = "free_y",
             labeller = labeller(sex = c(M = "Male", F = "Female"))) +
  scale_fill_manual(values = c(M = "#ff7f0e", F = "#1f77b4"), guide = "none") +
  labs(title = "Starting Salary Distribution by Sex", x = "Monthly Salary ($)", y = "Count") +
  theme_minimal()
```

## 3.4 Summary by Field and Sex

```{r}
aggregate(salary ~ field + sex, data = df,
          FUN = function(x) round(c(n = length(x), mean = mean(x), sd = sd(x)), 2))
```

```{r}
ggplot(df, aes(x = sex, y = salary, fill = sex)) +
  geom_boxplot(alpha = 0.7) +
  facet_wrap(~field) +
  scale_x_discrete(labels = c(M = "Male", F = "Female")) +
  scale_fill_manual(values = c(M = "#ff7f0e", F = "#1f77b4"), guide = "none") +
  labs(title = "Starting Salary by Sex and Field", x = NULL, y = "Monthly Salary ($)") +
  theme_minimal()
```

## 3.5 Salary by Hire Cohort

```{r}
cohort <- df %>%
  group_by(startyr, sex) %>%
  summarise(mean_sal = mean(salary), .groups = "drop")

ggplot(cohort, aes(x = startyr, y = mean_sal, colour = sex)) +
  geom_line() +
  geom_point() +
  scale_colour_manual(values = c(M = "#ff7f0e", F = "#1f77b4"),
                      labels = c(M = "Male", F = "Female")) +
  labs(title = "Mean Starting Salary by Hire Year and Sex",
       x = "Year Hired", y = "Mean Monthly Salary ($)", colour = "Sex") +
  theme_minimal()
```
Boxplots and histograms show the female distribution shifted lower than the male distribution (see figures on pages 3–4).
---

# 4. Inferential Analysis

## 4.1 Welch Two-Sample t-Test

$H_0: \mu_F = \mu_M$ vs $H_A: \mu_F \neq \mu_M$

```{r}
t.test(salary ~ sex, data = df, alternative = "two.sided")
```
Because the p-value is < 0.05, we reject the null hypothesis of equal means.

## 4.2 Multiple Linear Regression (Salary Level)

$$Y_i = \beta_0 + \beta_1\text{Female}_i + \beta_2\text{Field}_i + \beta_3\text{YrDeg}_i + \beta_4\text{HireYear}_i + \varepsilon_i$$

```{r}
ols <- lm(salary ~ sex + field + yrdeg + startyr, data = df)
coeftest(ols, vcov = vcovHC(ols, type = "HC3"))
```
This means that, holding those factors constant, women earn about $441 less per month at hire.

```{r}
summary(ols)$r.squared
```

## 4.3 Multiple Linear Regression (Log Salary)

Using log(salary) to interpret the sex gap as a percentage difference and to address right skew.

```{r}
log_ols <- lm(log(salary) ~ sex + field + yrdeg + startyr, data = df)
coeftest(log_ols, vcov = vcovHC(log_ols, type = "HC3"))
```

```{r}
summary(log_ols)$r.squared
```

The `sexF` coefficient from the log model gives the approximate percentage salary difference for women relative to men: `(exp(coef) - 1) * 100`. 

---

# 5. Diagnostics

```{r}
par(mfrow = c(2, 2))
plot(log_ols)
par(mfrow = c(1, 1))
```
What the analysis DOES show

- A large, statistically significant gender gap in starting salaries
- The gap persists even after controlling for several relevant factors
- Women earn roughly 9% less on average at hire

Yes, the results indicate a statistically significant gender difference in starting salaries, with women earning substantially less than men both before and after controlling for field, degree year, and hire year. The estimated gap is about $441 per month (~9%). Therefore, there is strong evidence of a gender salary disparity in starting pay. However, because observational data cannot establish causation, this analysis alone cannot definitively conclude that the disparity is due to wage discrimination.

---
