---
title: "Academic Field Analysis"
output: html_document
---

```{r setup, message=FALSE, warning=FALSE}
library(ggplot2)

salary <- read.table("salary.txt", header = TRUE)

salary$field <- as.factor(salary$field)
salary$sex <- as.factor(salary$sex)

field_palette <- c("#1b9e77", "#d95f02", "#7570b3")

base_theme <- theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = "right"
  )
```

# Academic Field: All Years

## Overall Distribution

Arts: **14.3%**  
Other: **66.4%**  
Prof: **19.2%**

The majority of faculty are in **Other** fields.

```{r}
field_counts_all <- table(salary$field)
field_props_all <- prop.table(field_counts_all)

field_counts_all
round(field_props_all, 3)
```

## Distribution Within Sex (Column Proportions)

| Field | Female | Male |
|-------|--------|------|
| Arts  | 20.4%  | 12.8% |
| Other | 69.7%  | 65.6% |
| Prof  | 9.9%   | 21.6% |

**Important Differences**

- Women are more concentrated in **Arts**
- Men are more concentrated in **Prof**
- **Other** is dominant for both sexes

```{r}
field_sex_counts_all <- table(salary$field, salary$sex)
field_sex_props_all <- prop.table(field_sex_counts_all, margin = 2)

field_sex_counts_all
round(field_sex_props_all, 3)
```

## Statistical Test

χ²(2) = **355.79**  
p < **2.2e−16**

→ Strong evidence that academic field and sex are associated.

```{r}
chisq.test(field_sex_counts_all)
```

## Visualizations

```{r fig.width=8, fig.height=6}
ggplot(salary, aes(x = field)) +
  geom_bar(fill = field_palette[1]) +
  labs(title = "Distribution of Academic Field (All Years)",
       x = "Academic Field",
       y = "Count") +
  base_theme
```

```{r fig.width=8, fig.height=6}
ggplot(salary, aes(x = sex, fill = field)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = field_palette) +
  labs(title = "Academic Field by Sex (All Years)",
       x = "Sex",
       y = "Proportion",
       fill = "Field") +
  base_theme
```

```{r fig.width=8, fig.height=6}
mosaicplot(field ~ sex,
           data = salary,
           main = "Mosaic Plot: Field by Sex (All Years)",
           col = field_palette)
```

---

# Academic Field — 1995 Only

```{r}
data95 <- subset(salary, year == 95)
```

## Overall Distribution

Arts: **13.8%**  
Other: **66.8%**  
Prof: **19.4%**

Very similar to the full dataset — good consistency.

```{r}
field_counts_95 <- table(data95$field)
field_props_95 <- prop.table(field_counts_95)

field_counts_95
round(field_props_95, 3)
```

## Distribution Within Sex (1995)

| Field | Female | Male |
|-------|--------|------|
| Arts  | 19.6%  | 11.8% |
| Other | 70.2%  | 65.7% |
| Prof  | 10.3%  | 22.6% |

**Differences Persist**

- Women are nearly twice as likely to be in **Arts** compared to men.
- Men are more than twice as likely to be in **Prof** compared to women.
- Distribution patterns remain stable over time.

```{r}
field_sex_counts_95 <- table(data95$field, data95$sex)
field_sex_props_95 <- prop.table(field_sex_counts_95, margin = 2)

field_sex_counts_95
round(field_sex_props_95, 3)
```

## Statistical Test

χ²(2) = **37.96**  
p = **5.73e−09**

→ Still statistically significant.

```{r}
chisq.test(field_sex_counts_95)
```

## Visualizations

```{r fig.width=8, fig.height=6}
ggplot(data95, aes(x = field)) +
  geom_bar(fill = field_palette[1]) +
  labs(title = "Distribution of Academic Field (1995)",
       x = "Academic Field",
       y = "Count") +
  base_theme
```

```{r fig.width=8, fig.height=6}
ggplot(data95, aes(x = sex, fill = field)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = field_palette) +
  labs(title = "Academic Field by Sex (1995)",
       x = "Sex",
       y = "Proportion",
       fill = "Field") +
  base_theme
```

```{r fig.width=8, fig.height=6}
mosaicplot(field ~ sex,
           data = data95,
           main = "Mosaic Plot: Field by Sex (1995)",
           col = field_palette)
```

---

# Overall Conclusion

Across all years and in 1995 specifically, academic field distribution is strongly associated with sex. Women are disproportionately represented in Arts, while men are disproportionately represented in Professional fields. These differences are statistically significant and remain stable over time, indicating persistent structural patterns in field representation.